pdp: An R Package for Creating Partial Dependence Plots

### 2016-09-04 manuscript/greenwell.tex
minor updates

\strong{Mention the integral approximation!!}

Let $\boldsymbol{x} = \left\{x_1, x_2, \dots, x_n\right\}$ represent the predictors in a model whose prediction function is $\widehat{f}\left(\boldsymbol{x}\right)$. If we partition $\boldsymbol{x}$ into an interest set, $\boldsymbol{z}_s$, and its compliment, $\boldsymbol{z}_{c} = \boldsymbol{x} \setminus \boldsymbol{z}_s$, then the partial dependence of the response on $\boldsymbol{z}_s$ is defined as
\begin{equation}
\label{eqn:pdf}
\bar{f}_s\left(\boldsymbol{z}_s\right) = \frac{1}{n}\sum_{i = 1}^n\widehat{f}\left(\boldsymbol{z}_s,\boldsymbol{z}_{i, c}\right),
\end{equation}
where $\boldsymbol{z}_{i, c}$ $\left(i = 1, 2, \dots, n\right)$ are the values of $\boldsymbol{z}_c$ that occur in the training sample; that is, we average out the effects of all the other predictors in the model. Let $x_1$ be the predictor variable of interest with unique values $\left\{x_{11}, x_{12}, \dots, x_{1k}\right\}$. The partial dependence of the response on $x_1$ can be constructed as follows:
\begin{enumerate}
  \item For $i \in \left\{1, 2, \dots, k\right\}$:
  \begin{enumerate}
    \item Create a copy of the training data and replace the original values of $x_1$ with the constant $x_{1i}$.
    \item Compute the vector of predicted values from the modified copy of the training data.
    \item Compute the average prediction to obtain $\bar{f}_1\left(x_{1i}\right)$.
  \end{enumerate}
  \item Plot the pairs $\left[x_{1i}, \bar{f}_1\left(x_{1i}\right)\right]$ for $i = 1, 2, \dotsc, k$.
\end{enumerate}
This can be quite computationally intensive since the algorithm involves $k$ passes over the training records. Fortunately, this can be parallelized quite easily (more on this in a later section). This algorithm can also be easily extended to larger subsets of 2 or more features as well.

@@

As described in the introduction, the \pkg{pdp} package is useful for constructing PDPs for many types of fitted models in R. These are especially useful for visualizing the relationships discovered by complex machine learning algorithms such as a random forest. The latest stable release is available from \href{https://cran.r-project.org/package=pdp}{CRAN}:
\begin{example}
install.packages("pdp")
\end{example}
The development version is located on GitHub: \url{https://github.com/bgreenwell/pdp}. Bug reports and usggestions are appreciated and should be sumbitted to \url{https://github.com/bgreenwell/pdp/issues}.

@@

\begin{table}[htbp]
  \begin{tabular}{p{4cm}ll}
    \toprule
      Type of model & R package & Object class \\
      \midrule
      Decision tree             & \CRANpkg{rpart} \citep{rpart-pkg} & \code{"rpart"} \\
                                & \pkg{party}    & \code{"BinaryTree"} \\
                                & \pkg{partykit} & \code{"constparty"}/\code{"party"} \\
      Bagged decision trees     & \CRANpkg{ipred} \citep{ipred-pkg} & \code{"classbagg"}, \code{"regbagg"} \\
                                & \CRANpkg{adabag} \citep{adabag-pkg} & \code{"bagging"} \\
      Boosted decision trees    & \pkg{gbm}      & \code{"gbm"} \\
                                & \CRANpkg{xgboost} & \code{"xgb.Booster"} \\
                                & \CRANpkg{adabag} \citep{adabag-pkg} & \code{"boosting"} \\
      Random forest & \pkg{randomForest} & \code{"randomForest"} \\
      Conditional random forest & \pkg{party}    & \code{"RandomForest"} \\
                                & \pkg{partykit} & \code{"cforest"} \\
      Linear model              & \pkg{stats}    & \code{"lm"} \\
      Generalized linear model  & \pkg{stats}    & \code{"glm"}, \code{"lm"} \\
      Nonlinear least squares   & \pkg{stats}    & \code{"nls"} \\
      Multivariate adaptive regression splines (MARS) & \CRANpkg{earth} \citep{earth-pkg} & \code{"earth"} \\
      Support vector machine    & \CRANpkg{e1071} \citep{e1071-pkg} & \code{"svm"} \\
                                & \CRANpkg{kernlab} \citep{kernlab-pkg} & \code{"ksvm"} \\
      \bottomrule
  \end{tabular}
  \caption{Models specifically supported by the \pkg{package}. \strong{Note:} package \pkg{xgboost} is only partially supported; see Section~\ref{sec:xgboost} fr details.}
  \label{tab:models}
\end{table}

@@

Next, we plot the partial dependence of \code{Species} on both \code{Petal.Width} and \code{Petal.Length} for each of the three classes. The result is displayed in Figure~\ref{fig:partial_iris}. Note that we had to supply the original training data to \code{partial} through the \code{train} option. This is necessary for fitted model objects that don't store a copy of the training data (e.g., model fitting functions that don't rely on a formula method).
\begin{example}
pd <- NULL
for (i in 1:3) {
  tmp <- partial(iris.svm, pred.var = c("Petal.Width", "Petal.Length"),
                 which.class = i, train = iris)
  pd <- rbind(pd, cbind(tmp, Species = levels(iris$Species)[i]))
}
lattice::levelplot(y ~ Petal.Width * Petal.Length | Species, data = pd)
\end{example}

>>>>

Let $\boldsymbol{x} = \left\{x_1, x_2, \dots, x_n\right\}$ represent the predictors in a model whose prediction function is $\widehat{f}\left(\boldsymbol{x}\right)$. If we partition $\boldsymbol{x}$ into an interest set, $\boldsymbol{z}_s$, and its compliment, $\boldsymbol{z}_c = \boldsymbol{x} \setminus \boldsymbol{z}_s$, then the partial dependence of the response on $\boldsymbol{z}_s$ is defined as
\begin{equation}
\label{eqn:pdf}
\bar{f}_s\left(\boldsymbol{z}_s\right) = \frac{1}{n}\sum_{i = 1}^n\widehat{f}\left(\boldsymbol{z}_s,\boldsymbol{z}_{i, c}\right),
\end{equation}
where $\boldsymbol{z}_{i, c}$ $\left(i = 1, 2, \dots, n\right)$ are the values of $\boldsymbol{z}_c$ that occur in the training sample; that is, we average out the effects of all the other predictors in the model. Equation~\eqref{eqn:pdf} approximates the average function
\begin{equation*}
  E_{\boldsymbol{z}_c}\left[\widehat{f}\left(\boldsymbol{z}_s\right)\right] = \int \widehat{f}\left(\boldsymbol{z}_s, \boldsymbol{z}_c\right)p_{c}\left(\boldsymbol{z}_c\right)d\boldsymbol{z}_c,
\end{equation*}
where $p_{c}\left(\boldsymbol{z}_c\right)$ is the marginal probability density of $\boldsymbol{z}_c$: $p_{c}\left(\boldsymbol{z}_c\right) = \int p\left(\boldsymbol{x}\right)d\boldsymbol{z}_s$.
%\begin{equation*}
%  p_{c}\left(\boldsymbol{z}_c\right) = \int p\left(\boldsymbol{x}\right)d\boldsymbol{z}_s.
%\end{equation*}

To construct a PDP in practice, we can proceed as follows. To simplify, let $\boldsymbol{z}_s = x_1$ be the predictor variable of interest with unique values $\left\{x_{11}, x_{12}, \dots, x_{1k}\right\}$. The partial dependence of the response on $x_1$ can be constructed as follows:
\begin{enumerate}
  \item For $i \in \left\{1, 2, \dots, k\right\}$:
  \begin{enumerate}
    \item Create a copy of the training data and replace the original values of $x_1$ with the constant $x_{1i}$.
    \item Compute the vector of predicted values from the modified copy of the training data.
    \item Compute the average prediction to obtain $\bar{f}_1\left(x_{1i}\right)$.
  \end{enumerate}
  \item Plot the pairs $\left[x_{1i}, \bar{f}_1\left(x_{1i}\right)\right]$ for $i = 1, 2, \dotsc, k$.
\end{enumerate}
This can be quite computationally intensive since the algorithm involves $k$ passes over the training records. Fortunately, this can be parallelized quite easily (more on this in a later section). This algorithm can also be easily extended to larger subsets of 2 or more features as well.

@@

As described in the introduction, the \pkg{pdp} package is useful for constructing PDPs for many types of fitted models in R. These are especially useful for visualizing the relationships discovered by complex machine learning algorithms such as a random forest. The latest stable release is available from \href{https://cran.r-project.org/package=pdp}{CRAN}:
\begin{example}
install.packages("pdp")
\end{example}
The development version is located on GitHub: \url{https://github.com/bgreenwell/pdp}. Bug reports and suggestions are appreciated and should be sumbitted to \url{https://github.com/bgreenwell/pdp/issues}.

@@

\begin{table}[htbp]
  \begin{tabular}{p{4cm}ll}
    \toprule
      Type of model & R package & Object class \\
      \midrule
      Decision tree             & \pkg{party}    & \code{"BinaryTree"} \\
                                & \pkg{partykit} & \code{"constparty"}/\code{"party"} \\
                                & \CRANpkg{rpart} \citep{rpart-pkg} & \code{"rpart"} \\
      Bagged decision trees     & \CRANpkg{adabag} \citep{adabag-pkg} & \code{"bagging"} \\
                                & \CRANpkg{ipred} \citep{ipred-pkg} & \code{"classbagg"}, \code{"regbagg"} \\
      Boosted decision trees    & \CRANpkg{adabag} \citep{adabag-pkg} & \code{"boosting"} \\
                                & \pkg{gbm}      & \code{"gbm"} \\
                                & \CRANpkg{xgboost} & \code{"xgb.Booster"} \\
      Random forest & \pkg{randomForest} & \code{"randomForest"} \\
      Conditional random forest & \pkg{party}    & \code{"RandomForest"} \\
                                & \pkg{partykit} & \code{"cforest"} \\
      Linear model              & \pkg{stats}    & \code{"lm"} \\
      Generalized linear model  & \pkg{stats}    & \code{"glm"}, \code{"lm"} \\
      Nonlinear least squares   & \pkg{stats}    & \code{"nls"} \\
      Multivariate adaptive regression splines (MARS) & \CRANpkg{earth} \citep{earth-pkg} & \code{"earth"} \\
      Support vector machine    & \CRANpkg{e1071} \citep{e1071-pkg} & \code{"svm"} \\
                                & \CRANpkg{kernlab} \citep{kernlab-pkg} & \code{"ksvm"} \\
      \bottomrule
  \end{tabular}
  \caption{Models specifically supported by the \pkg{pdp} package. \strong{Note:} package \pkg{xgboost} is only partially supported; see Section~\ref{sec:xgboost} for details.}
  \label{tab:models}
\end{table}

@@

Next, we plot the partial dependence of \code{Species} on both \code{Petal.Width} and \code{Petal.Length} for each of the three classes. The result is displayed in Figure~\ref{fig:partial_iris}. Note that we had to supply the original training data to \code{partial} through the \code{train} option. This is necessary for fitted model objects that don't store a copy of the training data (e.g., model fitting functions that don't rely on a formula method).
\begin{example}
pd <- NULL
for (i in 1:3) {
  tmp <- partial(iris.svm, pred.var = c("Petal.Width", "Petal.Length"),
                 which.class = i, .progress = "text",
                 grid.resolution = 101)
  pd <- rbind(pd, cbind(tmp, Species = levels(iris$Species)[i]))
}
lattice::levelplot(y ~ Petal.Width * Petal.Length | Species, data = pd)
\end{example}
